Climate indicators

### 2023-03-03 DRAFT_climate_indicators.Rmd
include spatial sd

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
#library(tidyverse) # cant update to v2 on server
library(dplyr)
library(ggplot2)
library(stars)
library(sf)
library(tmap)
library(lubridate)
library(ggpubr)
library(microbenchmark)
library(tictoc)
library(ncmeta)
library(parallel)
library(readr)
library(exactextractr)
library(tidyr)
library(terra)
dir <- substr(getwd(), 1,2)

@@

```{r, eval=F}
(
regional_means <- rast(y_var_norm) %>%
  exact_extract(reg, fun = 'mean', append_cols = "region", progress=T) %>%
  setNames(c("region", names(y_var_norm)))
  )
```

That was fast!

@@

```{r, eval=F}
(
regional_means_sd <- rast(y_var_norm) %>%
  exact_extract(reg, fun = function(X) c(mean = mean(x), sd = sd(x)), append_cols = "region", progress=T) %>%
  setNames(c("region", names(y_var_norm)))
  )
```

```{r, eval=F}
saveRDS(regional_means, "temp/regional_means.rds")
```

```{r}
regional_means <- readRDS("temp/regional_means.rds")
```

@@

(
regional_means_long <- regional_means %>%
  as.data.frame() %>%
  select(!all_of(div)) %>%
  pivot_longer(!region) %>%
  separate(name, into=c("type", "year")) %>%
  pivot_wider(#id_cols = region,
              names_from = type)  %>%
  left_join(temp, by=join_by(region)) %>% 
  mutate(diff = v-reference_upper) %>%
  mutate(threshold_low_centered = threshold_low-reference_upper) %>%
  mutate(threshold_high_centered = threshold_high-reference_upper)
)

@@

```{r}
(
i_2022 <- regional_means_long %>%
  filter(year >2017) %>%
  group_by(region) %>%
  summarise(i_2022 = mean(i))
)
```

```{r}
(
sd_2022 <- regional_means_long %>%
  filter(year >2017) %>%
  group_by(region) %>%
  summarise(i_2022 = mean(i))  
)
```

>>>>

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
#library(tidyverse) # cant update to v2 on server
library(dplyr)
library(ggplot2)
library(stars)
library(sf)
library(tmap)
library(lubridate)
library(ggpubr)
library(microbenchmark)
library(tictoc)
library(ncmeta)
library(parallel)
library(readr)
library(exactextractr)
library(tidyr)
library(data.table)
library(terra)
dir <- substr(getwd(), 1,2)

@@

```{r, eval=F}
system.time(
regional_means <- rast(y_var_norm) %>%
  exact_extract(reg, fun = 'mean', append_cols = "region", progress=T) %>%
  setNames(c("region", names(y_var_norm)))
  )
```

 user  system elapsed 
  7.603   3.071  10.697

That was fast! 
```{r}
regional_means
```

@@

```{r, eval=F}
system.time(
regional_sd <- rast(y_var_norm) %>%
  exact_extract(reg, fun = 'stdev', append_cols = "region", progress=T) %>%
  setNames(c("region", names(y_var_norm)))
  )
```

  user  system elapsed 
  7.420   2.921  10.355

```{r, eval=F}
saveRDS(regional_means, "temp/regional_means.rds")
saveRDS(regional_sd, "temp/regional_sd.rds")
```

```{r, include=F}
regional_means <- readRDS("temp/regional_means.rds")
regional_sd <- readRDS("temp/regional_sd.rds")
```

@@

regional_means_long <- regional_means %>%
  as.data.frame() %>%
  select(!all_of(div)) %>%
  pivot_longer(!region) %>%
  separate(name, into=c("type", "year")) %>%
  pivot_wider(#id_cols = region,
              names_from = type)  %>%
  left_join(temp, by=join_by(region)) %>% 
  mutate(diff = v-reference_upper) %>%
  mutate(threshold_low_centered = threshold_low-reference_upper) %>%
  mutate(threshold_high_centered = threshold_high-reference_upper)

#Adding the spatial sd
(
regional_means_long <- regional_sd %>%
  select(!all_of(div)) %>%
  pivot_longer(!region) %>%
  separate(name, into=c("type", "year")) %>%
  pivot_wider(names_from = type) %>%
  rename(i_sd = i,
         v_sd = v) %>%
  left_join(regional_means_long, by=join_by(region, year))
)
```

@@

```{r}
(
i_aggregatedToPeriods <- regional_means_long %>%
  mutate(year = as.numeric(year)) %>%
  mutate(period = case_when(
    year %between% c(2018, 2022) ~ "2018-2022",
    year %between% c(2013, 2017) ~ "2013-2017",
    year %between% c(2008, 2012) ~ "2008-2012",
    year %between% c(2003, 2007) ~ "2003-2007",
    .default = "pre 2003"
  )) %>%
  mutate(period_rank = case_when(
   period == "2018-2022" ~ 5,
   period == "2013-2017" ~ 4,
   period == "2008-2012" ~ 3,
   period == "2003-2007" ~ 2,
    .default = 1
  )) %>%
  group_by(region, period, period_rank) %>%
  summarise(indicator = mean(i),
            spatial_sd = sqrt(sum(i_sd^2))/length(i_sd)
            #,
            #spatial_sd_mean = mean(i_sd),
            #spatial_sd2_max = max(i_sd),
            #spatial_sd2_length = length(i_sd)
            )
)
```

```{r, fig.cap="Scaled indicator values, aggregated over 5 year intervals. Errors represent spatial variation within regions and across years."}
labs <- unique(i_aggregatedToPeriods$period[order(i_aggregatedToPeriods$period_rank)])

i_aggregatedToPeriods %>%
  ggplot(aes(x = period_rank, 
             y = indicator,
             colour=region))+
  geom_line() +
  geom_point() +
  geom_errorbar(aes(ymin=indicator-spatial_sd, 
                    ymax=indicator+spatial_sd), 
                    width=.2,
                 position=position_dodge(0.2)) +
  theme_bw(base_size = 12)+
  scale_x_continuous(breaks = 1:5,
                     labels = labs)+
  labs(x = "", y = "indikatorverdi")
```

Finally, I can add these values to the sp object.

```{r}
reg2 <- reg %>%
  left_join(i_aggregatedToPeriods[i_aggregatedToPeriods$period_rank==5,], by=join_by(region))

```{r}
ggplot(reg2) + 
  geom_sf(fill = indicator) +
  geom_sf_label(aes(label = indicator))

```{r}
tm_shape(reg2)+
  tm_polygons(col=)
```
